``Interestingness'' in Mathematical Theory Exploration

### 2019-10-25 PhDThesis/background_theoryexploration.tex
Add and rejiggle content

(Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
of definitions \iffalse TODO: Alison Some tools generate their own definitions
\fi in some formal system (for example a programming language) and
automatically generating a set of formal statements (properties) involving
those definitions. These may be conjectures or theorems (proven either by
sending conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps from a set of axioms). These
statements should also be
``interesting'' in some way, to rule out unhelpful trivialities such as
iterating a pattern over and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x},
\dots).

\iffalse
It is beyond the scope of this paper
to define what makes a mathematical statement ``interesting'', but this problem
has been tackled extensively in the literature~\cite{colton2000notion}.
\fi
% END TOTO

@@

The method of conjecture generation is a key characteristic of any theory
exploration system, although all existing implementations rely on brute force
enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
conjectures equations about Haskell code. We have used version 1 of \quickspec{}
in our experiments, due to its availability, stability and tooling integration;
hence references to \quickspec{} are to version 1 unless stated otherwise. At
the time of writing \iffalse TODO: Alison: Include date \fi there is a
\quickspec{} version 2 available, which has a much improved generation procedure
which can be significantly faster than its predecessor. Whilst \quickspec{} 2
has advanced the state of the art in theory exploration, our preliminary
experience has found that it still suffers the scaling issues we identify in
this work\footnote{Not only did we experience failures with our own experiments,
  we also suffered memory exhaustion when trying to run some of the examples
  included with the \quickspec{} 2 source code.} Since \quickspec{} 2 has less
integration with the tooling we have used, we leave a more thorough analysis of
it (and related systems such as \speculate{}) for future work, pending the
necessary infrastructure changes that would require.

@@

\quickspec{} (version 1) is written in the Haskell programming language, and
works with definitions which are also written in Haskell. A thorough description
of Haskell, including its suitability for theory exploration, can be found
in \S~\ref{sec:haskell}. For illustrative purposes, some simple Haskell
definitions are shown in Figure~\ref{fig:haskellteexample}. \quickspec{} uses
the following procedure to generate conjectures, which are both plausible (due
to testing on many examples) and potentially interesting (due to being mutually irreducible):

@@

The elements of $terms_{\hs{Nat}}$ will be grouped into five classes, one each
for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and
\hs{Bool}. As the variables \hs{a}, \hs{b} and \hs{c} are instantiated to
various randomly-generated numbers, the differences between these terms will
be discovered and the equivalence classes will be divided, until eventually
the equations in Figure \ref{fig:qspecresult} are conjectured.

Although complete, this enumeration approach is wasteful: many terms are
unlikely to appear in theorems, which requires careful choice by the user of
what to include in the signature. Here we know that addition and multiplication
are closely related, and hence obey many algebraic laws; arbitrary definitions
from a typical software library or proof development are unlikely to have be
related so strongly.

@@

\iffalse
% TODO: Work this into the example

As an example, giving the definitions from Figure~\ref{fig:haskellteexample} to
\quickspec{}, along with suitable comparison functions, random data generators and
variables \hs{n :: Nat} and \hs{b :: Bool}, will give rise to an equivalence
class for each type \hs{Nat}, \hs{Bool}, \hs{Nat -> Nat}, \hs{Nat -> Bool} and
\hs{Bool -> Bool}. Testing will find unequal terms in some of these classes
(such as the \hs{Bool} class containing \hs{True}, \hs{False}, \hs{b},
\hs{not b}, \hs{odd n} and \hs{even n}, which are all mutually unequal), split
them up, and repeat until no more splits are found. Equations are then generated
between elements of non-singleton classes, such as $\{\text{\hs{odd n}},
\text{\hs{not (even n)}}, \text{\hs{not (not (odd n))}}, \ldots\}$, and reduced
to a minimal set (for example, discarding \hs{odd n == not (not (odd n))} as an
instance of the more general \hs{b == not (not b)}). The remaining equations
(including \hs{not (not b) == b} and \hs{not (odd n) == even n}, etc.) are
output, either for the user to peruse or for another system to process, like the
\hipspec{} automated theorem prover.

Note that we must be able to generate values of a type in order to include
variables of that type. Also, only equivalence classes whose element type can be
compared are split and conjectured about in this way; in particular, functions
(like \hs{odd} and \hs{even}) are first-class values, so they will be put in an
equivalence class, but they will not be subject to testing and comparison since
Haskell has no generic notion of function equality~\footnote{In
  fact \quickspec{} compares expressions using a total ordering rather than equality;
  this is even more restrictive than requiring a notion of equality, but reduces
  the number of required comparisons}. \quickspec{} can hence conjecture that two
functions are pointwise-equal (e.g. \hs{f x == g x}) but it cannot conjecture
that the functions themselves are equal (e.g. \hs{f == g}).
\fi

@@

% \section{Exploration in Theorem Proving}
% \label{sec:examples}

% Before exploring abstract definitions of interestingness, we can first consider
% some scenarios which arise during formal proof where we are forced to generate
% conjectures. An analysis of these situations, and the subsequent theorems they
% produce, will contribute towards an empirical justification for what is
% interesting (at least from a utilitarian point of view) and inform our later
% exploration of the literature.

% \subsection{Generalisation}

% \providecommand{\coq}[1]{\lstinline[language=ML]|#1|}

% When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
% which $S$ is a special case. Although it seems counterintuitive, a generalised
% statement can sometimes be \emph{easier} to prove than the original. This arises
% often in inductive proofs, since the specific obligations which arise in the
% proof may be incompatible with the available inductive hypotheses.

% However, we cannot blindly generalise \emph{all} obligations we encounter, since
% \emph{over-generalising} results in obligations which are so strong that they
% are unprovable, or even false. We must therefore rely on heuristics to guide the
% generation of generalised conjectures, and hence perform a kind of exploration.

% An informative example is given by Boyer and Moore of the associativity of
% multiplication in ACL2 \cite{boyer1983proof}:

% $$(x * y) * z = x * (y * z)$$

% During the course of the proof, the following obligation arises:

% \begin{equation}
%   \tag{conc3}
%   (y + (x * y)) * z = (y * z) + ((x * y) * z)
%   \label{eq:conc3}
% \end{equation}

% ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
% sub-term $x * y$ with a fresh variable $w$:

% \begin{equation}
%   \tag{conc4}
%   (y + w) * z = (y * z) + (w * z)
%   \label{eq:conc4}
% \end{equation}

% This generalised form is clearly the distributivity law for multiplication and
% addition, which can be proved separately to the original goal of
% associativity. It would not be controversial to claim that this distributivity
% law is interesting in its own right (relative to associativity, at least), in
% addition to its usefulness in making this proof go through.

% % TODO: Describe the ACL2 heuristics

% Generalisation also occurs frequently when reasoning about \emph{tail-recursive}
% definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can
%   be executed in constant space using a loop, whereas recursion in non-tail
%   positions may require a growing number of stack frames or nested closures. See
%   \S \ref{sec:auxiliarylemmas} for example definitions of each type.}

% \subsection{Analogy}

% One way to characterise the interestingness of a statement is by \emph{analogy}
% to existing interesting statements. By finding lemmas analogous to those of a
% different theory, we may be able to re-use tactics and other forms of
% meta-programming across both.

% Existing theory exploration systems have been successfully applied to this
% problem, however the use of pure exploration misses opportunities to
% \emph{focus} the search, since we know which lemmas are used in those theories
% where a technique succeeded. If we can find an analogy to map from such solved
% problems to our unsolved goal, we can infer the approximate form of the lemmas
% we require, and target these specifically.

% The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
% to solving a goal $G$ by making analogies via unsupervised clustering
% \cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are used in two
% ways:

% \begin{itemize}
% \item First, we use the cluster $C_G$ containing $G$ to identify analogous
%   theorems.

% \item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
%   $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
%   to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
%   the cluster $C_s$ containing $s$.
% \end{itemize}

% The running examples for demonstrating \textsc{ACL2(ml)} are equivalence
% theorems for tail-recursive and non-tail-recursive calculations, as well as the
% effect of repeating certain list operations:

% \begin{itemize}

%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.

%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.

%   \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.

%   \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.

%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.

%   \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).

% \end{itemize}

% \subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}

% One consideration when generating conjectures is the difference between
% theorems, lemmas, corollaries, etc. From a logical point of view, these are all
% equivalent, and hence most proof assistants do not distinguish between
% them. However, their \emph{intention} may be different: in a sense, theorems are
% the interesting results; whilst lemmas are useful results, required for proving
% the theorems.

% Some systems, like Coq, allow users to \emph{label} each statement as being a
% \coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
% the same. This shows us immediately that lemmas outnumber theorems; in the Coq
% standard library there are over five times as many lemmas as theorems
% \footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when
%   excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of
%   \coq{Lemma} in its \texttt{theories/} directory.}.

% % TODO: Analyse them

% % TODO: Theory exploration as lemma generation; give example from a HipSpec paper

% We can find a need for auxiliary lemmas, once again, in the context of
% tail-recursive functions. Consider proving the (pointwise) equality of the
% following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:

% \begin{coqblock}
% Inductive Nat : Set := Z : Nat
%                      | S : Nat -> Nat.

% Fixpoint plus      (n m : Nat) := match n with
%                                       | Z    => m
%                                       | S n' => S (plus n' m)
%                                   end.

% Fixpoint plus_tail (n m : Nat) := match n with
%                                       | Z    => m
%                                       | S n' => plus_tail n' (S m)
%                                   end.
% \end{coqblock}

% \begin{haskell}
% Haskell equivalent:

% plus :: Nat -> Nat -> Nat
% plus      n  Z    = n
% plus      n (S m) = S (plus n m)

% plus_tail :: Nat -> Nat -> Nat
% plus_tail n  Z    = n
% plus_tail n (S m) = plus_tail (S n) m
% \end{haskell}

% Both of these functions implement addition, but the \coq{plus_tail} variant is
% tail-recursive. However, if we want to \emph{prove} that the definitions are
% (pointwise) equal, we run into difficulties. In particular, when the inductive
% step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
% reasonable), we cannot make this go through using another inductive argument.

% \begin{coqblock}
% (* Solve equalities by beta-normalising both sides *)
% Ltac triv := try (simpl; reflexivity).

% (* Prove equivalence of plus and plus_tail *)
% Theorem equiv : forall n m, plus n m = plus_tail n m.
%   induction n; triv. (* Base case is trivial *)

%   (* Inductive case: plus (S n) m = plus_tail (S n) m *)
%   intro m.

%   (* Beta-reduce the right-hand-side (justification is trivial) *)
%   replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.

%   (* Use induction hypothesis to replace plus_tail with plus *)
%   rewrite <- (IHn (S m)).
% \end{coqblock}

% Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
% the equation we need:

% \begin{coqblock}
% IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
% \end{coqblock}

% Yet we cannot provide it with the argument it needs, as our original induction
% hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):

% \begin{coqblock}
% IHn : forall x, plus (S n') x = plus_tail (S n') x
% \end{coqblock}

% We are forced to abandon the proof, despite such a reasonable-looking
% intermediate goal.

% In fact, if we attempt to prove that goal \emph{separately}, we can use a
% straightforward argument by induction; even though it is actually
% \emph{stronger} due to the absence of the \coq{IHn} assumption. Using this
% separate result as a lemma, the pointwise equality is proven easily.

% \begin{coqblock}
% Lemma gen n m : plus (S n) m = plus n (S m).
%   induction n; triv. (* Base case is trivial *)

%   (* Move all S constructors outside *)
%   simpl. rewrite <- IHn. simpl.

%   (* Trivial *)
%   reflexivity.
% Defined.
% \end{coqblock}

% \begin{coqblock}
%   rewrite (gen n m).
%   reflexivity.
% Defined.
% \end{coqblock}

>>>>

(Automated) Theory Exploration (TE) is the task of taking a \emph{signature}
of definitions in some formal system (for example a programming language) and
automatically generating a set of formal statements (properties) involving
those definitions\footnote{Some tools, such as \isascheme{}, also generate their
  own definitions. This work does not focus on such functionality, treating it
  instead under the umbrella of \emph{Concept Formation}.}. These may be
conjectures or theorems (proven either by sending conjectures to an automated
theorem prover, or by having the generating procedure proceed in logically sound
steps from a set of axioms). These statements should also be ``interesting'' in
some way, to rule out unhelpful trivialities such as iterating a pattern over
and over (e.g. \hs{x + 0 = x}, \hs{(x + 0) + 0 = x}, \dots). We defer a more
thorough treatment of what counts as ``interesting'' to
\S~\ref{sec:interestingness}.

@@

The method of conjecture generation is a key characteristic of any theory
exploration system, although all existing implementations rely on brute force
enumeration to some degree. We focus on \quickspec{}~\cite{QuickSpec}, which
conjectures equations about Haskell code. We have used version 1 of \quickspec{}
in our experiments, due to its availability, stability and tooling integration;
hence references to \quickspec{} are to version 1 unless stated otherwise. As of
2019 there is a \quickspec{} version 2 available, which has a much improved
generation procedure which can be significantly faster than its predecessor.
Whilst \quickspec{} 2 has advanced the state of the art in theory exploration,
our preliminary experience has found that it still suffers the scaling issues we
identify in this work\footnote{Not only did we experience failures with our own
  experiments, we also suffered memory exhaustion when trying to run some of the
  examples included with the \quickspec{} 2 source code.} Since \quickspec{} 2
has less integration with the tooling we have used, we leave a more thorough
analysis of it (and related systems such as \speculate{}) for future work,
pending the necessary infrastructure changes that would require.

@@

\quickspec{} is written in the Haskell programming language, and works with
definitions which are also written in Haskell. A thorough description of
Haskell, including its suitability for theory exploration, can be found in
\S~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
are shown in Figure~\ref{fig:haskellteexample}. \quickspec{} uses the following
procedure to generate conjectures, which are both plausible (due to testing on
many examples) and potentially interesting (due to being mutually irreducible):

@@

The elements of $terms_{\hs{Nat}}$ will be grouped into five equivalence classes,
one each for \hs{Nat}, \hs{Nat -> Nat}, \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool}
and \hs{Bool}. As the variables \hs{a}, \hs{b} and \hs{c} are instantiated to
various randomly-generated numbers, the differences between these terms will
be discovered. For example, the \hs{Nat} class will initially contain (among
other things) the terms \hs{Z}, \hs{x} and \hs{plus x Z}; this will be split
apart once \hs{x} is tested with a non-zero value, since the latter terms will
not equal \hs{Z}. Since those latter terms will always be equal, for any value
of \hs{x}, they will never be split apart. Eventually (after removing
redundancies) the equations in Figure~\ref{fig:qspecresult} are conjectured.
These may be output to the user for perusal, or for another system to process
like the \hipspec{} automated theorem prover.

Note that we must be able to generate values of a type in order to include
variables of that type. Also, only equivalence classes whose element type can be
compared are split and conjectured about in this way\footnote{Both random data
  generators and comparison functions can be derived mechanically from algebraic
  datatype declarations; hence why they are omitted from our example. Note that
  care must be taken when generating recursive structures, as na\"ive recursion
  may produce values whose expected size exceeds the machine's memory.}; in
particular, functions (like \hs{odd} and \hs{even}) are first-class values, so
they will be put in an equivalence class, but they will not be subject to
testing and comparison since Haskell has no generic notion of function
equality~\footnote{In fact \quickspec{} compares expressions using a total
  ordering rather than equality; this is even more restrictive than requiring a
  notion of equality, but reduces the number of required comparisons}.
\quickspec{} can hence conjecture that two functions are pointwise-equal (e.g.
\hs{f x == g x}) but it cannot conjecture that the functions themselves are
equal (e.g. \hs{f == g}).

Although complete, this enumeration approach is wasteful: many terms are
unlikely to appear in theorems, which requires careful choice by the user of
what to include in the signature. Here we know that addition and multiplication
are closely related, and hence obey many algebraic laws; arbitrary definitions
from a typical software library or proof development are unlikely to be related
so strongly.

@@

@@

### 2019-10-25 PhDThesis/litreview_interestingness.tex
Fix reference
Add and rejiggle content

\subsection{Interestingness}

A major challenge when generating conjectures is choosing how to narrow down the
resulting set to only those deemed ``interesting'', since this is an imprecise
term with many different interpretations. For example, all existing approaches
agree that simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton \etal{} surveyed tools for theory formation
and exploration, and their associated notions of ``interestingness'' for
concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
which are applied to conjectured properties as follows:

@@

% FIXME: This is original work, so should be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{             &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti     & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}     & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table~\ref{table:colton} and state which were used
in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
based on this survey (the latter is Colton's own tool). We extend this analysis
to some more recent tools in the remaining rows (based on our understanding of
their function): \quickspec{}~\cite{QuickSpec},
\speculate{}~\cite{braquehais2017speculate},
\isacosy{}~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and \isascheme{}~\cite{MontanoRivas2011}.

@@

>>>>

\section{Interestingness}
\label{sec:interestingness}

A major challenge when generating conjectures is choosing how to focus attention
on only those deemed ``interesting'' (either by avoiding uninteresting search
paths, and/or by filtering down the eventual output), since this is an imprecise
term with many different interpretations. For example, all existing approaches
agree that simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton~\etal{} surveyed tools for theory formation
and exploration, and their associated notions of ``interestingness'', for
concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
which are applied to conjectured properties as follows:

@@

% FIXME: This is original work, so should maybe be in a different section
\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{             &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti     & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}     & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR           & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{\quickspec{} & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{\speculate{} & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{\isacosy{}   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{\isascheme{} & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table~\ref{table:colton} and state which were used
in key ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are
based on this survey (the latter is Colton's own tool). The remaining rows are
our extension of this analysis to the (higher-order) tools we have focused on
(based on our understanding of their function): \quickspec{}~\cite{QuickSpec},
\speculate{}~\cite{braquehais2017speculate},
\isacosy{}~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and \isascheme{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}.

@@

\subsection{Exploration in Theorem Proving}
\label{sec:examples}

Conjecture generation (automated or manual) is crucial when formally proving
statements, such verifying software against its specification. This provides a
concrete, empirical setting to determine what makes statements interesting (at
least from a utilitarian point of view).

\paragraph{Generalisation}

\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs (e.g. when reasoning about tail-recursive
definitions~\cite{kapur2003automatic}), since the specific obligations which
arise in the proof may be incompatible with the available inductive hypotheses.
An informative example is given by Boyer and Moore of the associativity of
multiplication in ACL2~\cite{boyer1983proof}:

$$(x * y) * z = x * (y * z)$$

During the course of the proof, the following obligation arises:

\begin{equation}
  \tag{conc3}
  (y + (x * y)) * z = (y * z) + ((x * y) * z)
  \label{eq:conc3}
\end{equation}

ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
sub-term $x * y$ with a fresh variable $w$:

\begin{equation}
  \tag{conc4}
  (y + w) * z = (y * z) + (w * z)
  \label{eq:conc4}
\end{equation}

This generalised form is clearly the distributivity law for multiplication and
addition, which can be proved separately to the original goal of
associativity. It would not be controversial to claim that this distributivity
law is interesting in its own right (relative to associativity, at least), in
addition to its usefulness in making this proof go through.

% TODO: Describe the ACL2 heuristics

Generality hence contributes to a conjecture's \emph{utility}, and also (by
definition) to its \emph{applicability}. It is still important to balance
against other concerns, since blindly generalising \emph{all} statements of
proof obligations we come across would result in conjectures so strong that they
are unprovable, or even false; which would make them much less interesting.

\paragraph{Analogy}

The interestingness of a statement can also be characterised by \emph{analogy}
to existing interesting statements. By finding lemmas analogous to those of a
different theory, we may be able to re-use proof tactics and other forms of
meta-programming across both.

Existing theory exploration systems have been successfully applied to this
problem, however the use of pure exploration misses opportunities to
\emph{focus} the search, since we know which lemmas are used in those theories
where a technique succeeded. If we can find an analogy to map from such solved
problems to a currently-unproven goal, we can infer the approximate form of the
lemmas we require, and target these specifically.

The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
to solving a goal $G$ by making analogies via unsupervised
clustering~\cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are
used in two ways:

\begin{itemize}
\item First, we use the cluster $C_G$ containing $G$ to identify analogous
  theorems.

\item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
  $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
  to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
  the cluster $C_s$ containing $s$.
\end{itemize}

The interestingness of such generated lemmas derives from both the existence of
the original lemma (which is presumably interesting enough to prove and
subsequently explore), that of the target theorem it may help to prove, and the
novelty provided by reframing existing concepts in a new domain. The running
examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
tail-recursive and non-tail-recursive calculations\footnote{A tail-recursive
  function can be executed in constant space using a loop, whereas recursion in
  non-tail positions may require a growing number of stack frames or nested
  closures.}), as well as the effect of repeating certain list operations:

\begin{itemize}

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.

  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.

  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.

  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).

\end{itemize}

\paragraph{Auxiliary Lemmas} \label{sec:auxiliarylemmas}

One consideration when generating conjectures is the difference between
theorems, lemmas, corollaries, etc. From a logical point of view, these are all
equivalent, and hence most proof assistants do not distinguish between
them. However, their \emph{intention} may be different: in a sense, theorems are
the interesting results; whilst lemmas are useful results, required for proving
the theorems. Corollaries are consequences and special-cases of theorems, which
are interesting enough to state separately, but not as much as a theorem.

Some systems, like Coq, allow users to \emph{label} each statement as being a
\coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
the same. This shows us immediately that lemmas outnumber theorems; in the Coq
standard library there are over five times as many lemmas as theorems
\footnote{Coq version \texttt{8.4pl6} contains (excluding comments) 1492
  occurences of \coq{Theorem} and 7594 of \coq{Lemma} in its \texttt{theories/}
  directory.}.

% TODO: Analyse them

% TODO: Theory exploration as lemma generation; give example from a HipSpec paper

We can find a need for auxiliary lemmas, once again, in the context of
tail-recursive functions. Consider proving the (pointwise) equality of the
following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:

\begin{coqblock}
Inductive Nat : Set := Z : Nat
                     | S : Nat -> Nat.

Fixpoint plus      (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => S (plus n' m)
                                  end.

Fixpoint plus_tail (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => plus_tail n' (S m)
                                  end.
\end{coqblock}

These definitions are equivalent to the following Haskell:

\begin{haskell}
plus :: Nat -> Nat -> Nat
plus      n  Z    = n
plus      n (S m) = S (plus n m)

plus_tail :: Nat -> Nat -> Nat
plus_tail n  Z    = n
plus_tail n (S m) = plus_tail (S n) m
\end{haskell}

Both of these functions implement addition, but the \coq{plus_tail} variant is
tail-recursive. However, if we want to \emph{prove} that the definitions are
(pointwise) equal, we run into difficulties. In particular, when the inductive
step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
reasonable), we cannot make this go through using another inductive argument.

\begin{coqblock}
(* Solve equalities by beta-normalising both sides *)
Ltac triv := try (simpl; reflexivity).

(* Prove equivalence of plus and plus_tail *)
Theorem equiv : forall n m, plus n m = plus_tail n m.
  induction n; triv. (* Base case is trivial *)

  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
  intro m.

  (* Beta-reduce the right-hand-side (justification is trivial) *)
  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.

  (* Use induction hypothesis to replace plus_tail with plus *)
  rewrite <- (IHn (S m)).
\end{coqblock}

Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
the equation we need:

\begin{coqblock}
IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
\end{coqblock}

Yet we cannot provide it with the argument it needs, as our original induction
hypothesis is \emph{too specific} (it has too many \coq{S} constructors):

\begin{coqblock}
IHn : forall x, plus (S n') x = plus_tail (S n') x
\end{coqblock}

We are forced to abandon the proof, despite such a reasonable-looking
intermediate goal.

In fact, if we attempt to prove that goal \emph{separately}, we can use a
straightforward argument by induction; even though it is actually
\emph{stronger} due to the absence of the \coq{IHn} assumption:

\begin{coqblock}
Lemma gen n m : plus (S n) m = plus n (S m).
  induction n; triv. (* Base case is trivial *)

  (* Move all S constructors outside *)
  simpl. rewrite <- IHn. simpl.

  (* Trivial *)
  reflexivity.
Defined.
\end{coqblock}

Using this separate result as a lemma, the orignal pointwise equality is proven
easily:

\begin{coqblock}
  rewrite (gen n m).
  reflexivity.
Defined.
\end{coqblock}

Hence demonstrating the utility, and therefore interestingness, of such
auxilliary lemmas.
